Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-26 Manuscript/manuscript.Rmd
edits to methods

## Computational framework for boundary prediction

@@

Cell type-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

Chromosome-specific datasets used for modeling were composed of two components: the response vector $Y_{i}$, an indicator variable distinguishing boundary and non-boundary regions, and the predictor-type specific data matrix, $\textbf{X}_{ij}$, where $i \in \{ 1,2, \cdots,N \}$ ($N$ being the number of chromosome-specific bins) and $j \in \{ 1,2,\cdots,P \}$ ($P$ being the number of genomic annotations). <!--Chromosome-specific datasets were concatenated to form a whole genome dataset. The dataset was then split into a 7:3 ratio for training and testing, respectively.[!!!only comes into play with preciseTAD]-->

## Class imbalance assessment

@@

## Assessment of model performance

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.-->

@@

## Boundary prediction at bp resolution

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin  loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements) [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. The *nodesize* and *ntree* parameters were tuned sequentially using a grid search across values from ???-??? by ??? and from ???-??? by ??? for each, respectively. We then used the same optimized whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). *preciseTAD* guards against inflating the number of predicted boundaries by combining both density-based spatial clustering (DBSCAN) and partitioning techniques to identify precise TAD (or loop) boundaries throughout the linear genome. First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances [??? Is that true? Matrix is 2-dimensional] between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution [??? Meaning?]. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3 [??? Meaning?]. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR) [??? Use simple words: "boundary region" (many bases clustered by DBSCAN), then "boundary base" (selected by PAM)]. To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.

<!--
We then performed hierarchical clustering on the distance matrix (in base pairs) constructed from the midpoints of each PTBA using complete linkage clustering. To identify the optimal number of clusters, *k*, we calculated the mean silhouette score over all possible sizes of clusters *m*:

$$\bar{s}_{m} = \dfrac{\sum^{m}_{i} s_{i}}{m},$$

where $s_{i}$ is the cluster-specific silhouette score defined as,

$$s_{i} = \dfrac{b_{i} - a_{i}}{max[a_{i},b_{i}]}.$$

Here, $a_{i}$ is the mean distance between each cluster entry and all other observations in the same cluster, and $b_{i}$ is the mean distance between entries within the same cluster. Therefore number of clusters, *k*, that maximized the mean silhouette score, $\bar{s}_{m}$ were used to identify groupings of PTBA, which we defined as *predicted TAD boundary regions* (PTBR).
-->

<!--
Using optimally trained models, we established *predicted TAD boundary regions* as a collection of adjacent bases predicted as TAD boundary with the probability of 1. [??? Do not understand from here:] To guard against inflating the number of predicted TAD boundaries, we clustered the distances between the midpoints of the *predicted TAD boundary regions*. We then performed partitioning around mediods (PAM) to determine the *predicted TAD boundary region* that was most representative of the cluster. The midpoint, in this case a particular basepair, of the *predicted TAD boundary region* identified as the medoid was defined as a final *predicted TAD boundary*. [??? To here. REWRITE!]
-->

@@

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell types and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, the set of shared boundaries divided by the total number of unique boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. The alpha level of significance was set to 0.05 for all statistical tests.

>>>>

## Shifted-binning for binary classification

@@

## Feature engineering

Cell line-specific genomic annotations were used to build the feature space for our boundary prediction framework. Bins were annotated by one of either the number of overlaps with an annotation (*Overlap Count (OC)*), the percent of overlap between the bin and the total width of genomic annotation regions overlapping it (*Overlap Percent (OP)*), the average signal strength of the corresponding annotation (*Peak Signal Strength*), or the distance in bases from the center of the bin to the center of the nearest genomic annotation region (*Distance*). A $log_2$-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). It should be noted that models built using a *Peak Signal Strength* feature space were only composed of histone modifications and transcription factor binding sites. That is, they did not include BroadHMM chromatin segmentations.

## Addressing class imbalance

@@

## Establishing computational framework for assessing model performance

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. First, the data matrix $A_{N \times (p+1)}$ was constructed by combining the binned genome from each of the $n-1$ autosomal chromosomes ($1,2,\cdots, i-1,i+1,\cdots,22$) as well as a particular feature engineering procedure. Re-sampling was then performed on $A$. An RF classifier was then trained on $A$ using 5-fold cross-validation to tune for the number of annotations, *p*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. To evaluate model performance, the $i^{th}$ chromosome was used for testing. The performance of each combination of model inputs was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. <!--We refer to the complete computational framework used to compare random forest model inputs as *TADRF*.--> Performances were aggregated across each $i^{th}$ holdout chromosome using means and standard deviations.

@@

## Boundary prediction at bp resolution using *preciseTAD*

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we implemented an optimally built RF model using the framework described above. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements).

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The eps parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair within each clustered PTBR.

@@

Furthermore, under the hypothesis that TAD boundaries are highly concerved across cell types [@dixon2012topological; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa], we assessed the conservation of our predicted TAD boundaries across cell lines and prediction algorithms. Boundaries were flanked by resolution and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0) and evaluated using the Jaccard index defined as

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, between cell lines, the number of overlapping flanked boundaries divided by the total number of flanked boundaries. Wilcoxon rank sum tests were used to compare chromosome-specific jaccard indices across cell lines, between *preciseTAD* boundaries and both ARROWHEAD and PEAKACHU boundaries respectively. The alpha level of significance was set to 0.05 for all statistical tests.
